### 2014-05-10 writeups/context-dependence.tex
Minor changes.

The common framework of the above examples is a one-dimensional grid of $n$ sites,
and each site is labeled with one of a finite set of states.
These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
where each $X_i(t) \in \calS$, the set of possible states.
(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
We define a \textit{pattern} to mean a contiguous sequence of states, and we let $|u|$ denote the length of the pattern $u$.
The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
Thus given two patterns $u$ and $v$ of common length $h$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+h}(t)) = (u_1, \ldots, u_h)$ then
\[
\P\{(X_i(t+dt), \ldots, X_{i+h}(t+dt)) = (v_1, \ldots, v_h) = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
We call such a $(\mu,u,v)$ a ``transition triple''.
The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$
determines the time evolution of the process.

@@

\paragraph{The generator matrix}
For clarity, we should define how the set of transition rates determines the transition rate matrix,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x$ jumps to state $y$.
% Let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ be the subsequence of length $h$ beginning at location $i$, and
For each $1\le i \le n$, pattern length $h \ge 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h, and \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+h-1$, where $x$ matches with $u$ and $y$ matches with $v$.
(Interpet these indices modulo $n$, even for short sequences, a point we will come back to.)
We'll also need the set of defined transitions in $\calT$
that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
that could change $x$ into $y$.
In principle, this gives us the transition probabilities for the process:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
but in practice the size of these matrices can make a direct application of this formula inefficient.
This paper proposes an alternate strategy.

@@

However, in any case there is no guarantee that this is close to correct,
as it ignores dependencies between neighboring blocks.
Intuitively, we want to increase the size of the context
-- but increasing $m$ does not do this, as there are always edge effects.
\todo[inline]{I'm a little confused by this use of "edge effects." Are you using it to mean non-independence between blocks generally?}
However, we could look at the context in the initial sequence but not in the final sequence,
including the states of an additional $\ell$ and $r$ sites on the left and right, respectively.
We will call these left and right \emph{fringes}.
Concretely, define the marginal probabilities
\begin{align}
  p_{\ell,m,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_{\ell+m+r}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{\ell+m+r}, \; y \in \calS^m ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
This is the probability that if we begin with a sequence $x$ of length $\ell+m+r$,
then after time $t$ we will see at the middle $m$ positions the pattern $y$,
ignoring what is present in the leftmost $\ell$ and rightmost $r$ positions.
In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z | X_i(0)^{(m)} = x \}$,
we do expect $p_{\ell,m,r}(x,y)$ to asymptotically correct when $\ell$ and $r$ are large:
\begin{align} \label{eqn:window_approx}
  \P\{ X_i^{(m)}(t) = y | X_i(0)^{(\ell+m+r)} = x \} \approx p_{\ell,m,r}(x,y) .
\end{align}
We provide a precise statement of this approximation later.

@@

Now we will now calculate the mean and covariances of $N_\ell(x,y)$.
From the definitions above,
the expected number of times $y$ would appear juxtaposed with $x$ is approximately $N(x) p_{\lyl}(x,y)$.
Covariances are also easy to compute.
If we define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$, then
\begin{align}
  \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^n \cov[M_1(x,y),M_{k+1}(x,y)] \\
  &\approx n p_{\lyl}(x,y)(1-p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2 \right) \\
  &= n p_{\lyl}(x,y)(1-(1+2m+4\ell)p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
\end{align}
The term $\E[M_1(x,y) M_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
\begin{align}
  \begin{split}
    \cov[N(x_1,y_1),N(x_2,y_2)]
    & \approx - n (2m+4\ell) p_{\lyl}(x_1,y_1)p_{\lyl}(x_2,y_2) \\
    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
  \end{split}
\end{align}

>>>>

The common framework of the above examples is a one-dimensional grid of $n$ sites,
and each site is labeled with one of a finite set of states.
These change with time -- so, write $X_i(t)$ for the state of the $i^\mathrm{th}$ site at time $t$,
where each $X_i(t) \in \calS$, the set of possible states.
(To avoid edge effects, we sometimes arrange these sites in a circle, so the indices are interpreted $\mod n$.)
We define a \textit{pattern} to mean a contiguous sequence of states, and we let $|u|$ denote the length of the pattern $u$.

The marginal mutation process at every context is continuous time homogeneous Markov, which are thus specified by a set of transition rates between patterns.
The sequence-level transition process is the independent superposition of all of the individual pattern mutation processes.
Thus given two patterns $u$ and $v$ of common length $h$, saying ``pattern $u$ changes to $v$ at rate $\mu$'' means that if $(X_i(t), \ldots, X_{i+h}(t)) = (u_1, \ldots, u_h)$ then
\[
\P\{(X_i(t+dt), \ldots, X_{i+h}(t+dt)) = (v_1, \ldots, v_h) = \mu dt + o(dt),
\]
regardless of the location $i$ along the sequence.
If more than one pattern would match to cause the same change, their rates add.
We call such a $(\mu,u,v)$ a ``transition triple''.
The set of all transition triples $\calT = \{ (\mu^j,u^j,v^j) \}_{j=1}^{n_T}$
determines the time evolution of the process.

@@

\paragraph{The generator matrix}
For clarity, we should define how the set of transition rates determines the transition rate matrix,
the $|\calS|^n \times |\calS|^n$ matrix $G(n)$ whose $(x,y)^\text{th}$ entry gives the instantaneous rate
with which the process in state $x$ jumps to state $y$.
Let $x_i^{(h)} = (x_i, x_{i+1}, \ldots, x_{i+h-1})$ be the subsequence of length $h$ beginning at location $i$.
For each $1\le i \le n$, pattern length $h \ge 0$, and patterns $u,v \in \calS^h$ define the relation
\[
x \xrightarrow{i,u,v} y \qquad \text{iff} \qquad \begin{cases}
  x_k = y_k \quad &\text{for } k<i \\
  x_{i+k} = u_k \quad &\text{for } 0 \le k < h \\
  y_{i+k} = v_k \quad &\text{for } 0 \le k < h, and \\
  x_k = y_k \quad &\text{for } k\ge i+h ,
\end{cases}
\]
i.e.\ if $x$ and $y$ match except for in $i,i+1,\ldots,i+h-1$, where $x$ matches with $u$ and $y$ matches with $v$.
(Interpet these indices modulo $n$, even for short sequences, a point we will come back to.)
We'll also need the set of defined transitions in $\calT$
that can move from $x$ to $y$ at a given location: let  $J(i,x,y) = \{ (\mu^j,u^j,v^j) \in \calT \st x \xrightarrow{i,u^j,v^j} y \}$.
Then, the rate $G(n)_{x,y}$ is the sum of all matching transition rates,
namely
\begin{align} \label{eqn:G_defn}
  G(n)_{x,y} = \sum_{i=1}^n \sum_{J(i,x,y)}  \mu^j ,
\end{align}
and if there are no triples $(\mu,u,v)$ with $x \xrightarrow{i,u,v} y$ for some $i$, then $G(n)_{x,y}=0$.
This is just to say that the transition rate from $x$ to $y$ is the sum of the mutation rates of all transition triples
that could change $x$ into $y$.
In principle, this gives us the transition probabilities for the process:
\begin{align} \label{eqn:full_likelihood}
    p_n(t;x,y) := \P\{ X(t) = y \mid X(0) = x \} = \left(e^{tG(n)}\right)_{xy} ,
\end{align}
but in practice the size of these matrices can make a direct application of this formula inefficient.
This paper proposes an alternate strategy.

@@

However, in any case there is no guarantee that this is close to correct,
as it ignores dependencies between neighboring blocks.
Intuitively, we want to increase the size of the context
-- but increasing $m$ does not do this, as there are always edge effects.
\todo[inline]{I'm a little confused by this use of "edge effects." Are you using it to mean non-independence between blocks generally?}
However, we could look at the context in the initial sequence but not in the final sequence,
including the states of an additional $\ell$ and $r$ sites on the left and right, respectively.
We will call these left and right \emph{fringes}.
Concretely, define the marginal probabilities
\begin{align}
  p_{\ell,m,r}(t;x,y) &= \sum_{a \in \calS^\ell} \sum_{b \in \calS^r} p_{\ell+m+r}(t;x,a \join y \join b) , \quad \text{for}\; x \in \calS^{\ell+m+r}, \; y \in \calS^m ,
\end{align}
where $a \join y \join b$ is the sequence composed of concatenating $a$, $y$, and $b$ together in that order.
This is the probability that if we begin with a sequence $x$ of length $\ell+m+r$,
then after time $t$ we will see at the middle $m$ positions the pattern $y$,
ignoring what is present in the leftmost $\ell$ and rightmost $r$ positions.
\todo[inline]{Is the following notation right? It seems like taking it out is a viable option: "In a long sequence, $p_m(x,z)$ might not be a very good approximation for $\P\{ X_i^{(m)}(t) = z \mid X_i^{(m)}(0) = x \}$,"}
We expect $p_{\ell,m,r}(x,y)$ to be asymptotically correct when $\ell$ and $r$ are large \todo{and $t$ is not} :
\begin{align} \label{eqn:window_approx}
  p_{\ell,m,r}(x,y) \approx \P\{ X_i^{(m)}(t) = y \mid X_i^{(\ell+m+r)}(0) = x \}.
\end{align}
We provide a precise statement of this approximation later.

@@

Now we will now calculate the expectation and covariances of $N_\ell(x,y)$.
From the definitions above,
\[
\E [N_\ell(x,y)] \approx N(x) \, p_{\lyl}(x,y).
\]
To compute covariances, define $M_i(x,y)$ to be 1 if $X_{i-\ell}^{(|x|)}(0)=x$ and $X_i^{(|y|)}(t)=y$, and 0 otherwise,
so that $N(x,y) = \sum_{i=1}^n M_i(x,y)$.
\begin{align*}
  \var[N(x,y)] &= \sum_{i=1}^n \var[M_i(x,y)] + \sum_{i\neq j} \cov[M_i(x,y), M_j(x,y)] \\
  &= n \var[M_1(x,y)] + 2n\sum_{k=1}^{n-1} \cov[M_1(x,y),M_{k+1}(x,y)] \\
  &\approx n p_{\lyl}(x,y)(1-p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \left( \E[M_1(x,y) M_{k+1}(x,y)] - p_{\lyl}(x,y)^2 \right) \\
  &= n p_{\lyl}(x,y)(1-(1+2m+4\ell)p_{\lyl}(x,y)) + 2n\sum_{k=1}^{m+2\ell} \E[M_1(x,y) M_{k+1}(x,y)] .
\end{align}
The term $\E[M_1(x,y) M_{k+1}(x,y)]$ is the probability that $x$ matches at both site 1 and site $k+1$,
and that $y$ matches both at site $\ell+1$ and at site $\ell+k+1$.
For this to be nonzero, the joint pattern $(x,y)$ must allow self-overlaps.
Similarly, if $(x_1,y_1) \neq (x_2,y_2)$ then
\begin{align}
  \begin{split}
    \cov[N(x_1,y_1),N(x_2,y_2)]
    & \approx - n (2m+4\ell) p_{\lyl}(x_1,y_1)p_{\lyl}(x_2,y_2) \\
    & \qquad {} + n\sum_{k=1}^{m+2\ell} \E[M_1(x_1,y_1) M_{k+1}(x_2,y_2) + M_1(x_2,y_2) M_{k+1}(x_1,y_1)] .
  \end{split}
\end{align*}
